---
title: "Introduction to the ptest package"
author: "Yuanhao Lai, Ian McLeod"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ptest.bib
vignette: >
  %\VignetteIndexEntry{Introduction to the ptest package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

## Abstract

This vignette introduces the functionality of the `ptest` package for 
fitting harmonic models and detecting the periodicty for short time series. 

## Motivation

Currently there are many statistical tests available for detecting the periodicity in microarray gene expression data. 
However, methods like the Fisher’s g test proposed by [@Wichert01012004] might fail to detect the periodicity as it only consider the Fourier frequencies, and methods like the robust g test [@ahdesmaki2005robust] lacks of the computational efficiency as the test requires time-consuming Monte Carlo simulations. 

The figure below shows the gene expression measured at 11 hourly intervals for gene ORF06806, which is apparently periodic with period of about 8.4 hours. 
About 7 other genes (out of 3000 available) show strong periodicity that is not detected by Fisher's g or other tests in current use for detecting periodicity in time series microarrays [@Wichert01012004].

```{r Motivation, fig.cap="Gene ORF06806",fig.width=5,fig.height=3,echo=FALSE}
library(ptest)
data(Cc)
x <- Cc[which(rownames(Cc)=="ORF06806"),]

plot(1:length(x),x,type="b",
     xlab="time",ylab="Gene expression")
```

We introduce `ptest` as an efficient and elementary tool for modeling and testing periodicity in short time series, which aims to deal with the above obstacles. 
It includes serveral periodicity tests, whose p-value can be obtained instantly and accurately by the response surface regression (RSR) method [@mackinnon2000computing]. 
It also includes a likelihood ratio test based on the harmonic regression model with the estimated frequency searched among $\mathcal{E}_{n}=\{ j/101\mid j=1,\dots,50\wedge j/101\ge1/n\}$, which allows for detecting both Fourier frequencies and the non-Fourier frequencies.

## Harmonic model
For the introduced methods, we assume there are no missing values in the data and the model for the periodic time series is,
$$z_{t}=\mu+A\cos(2\pi\lambda t)+B\sin(2\pi\lambda t)+\varepsilon_{t},\ t=1,...,n,$$

where $\epsilon_{t}$ is an i.i.d. noise sequence with mean 0 and
variance $\sigma^{2}$, the frequency $\lambda$ of the sinusoidal
wave is restricted to the range $0\le\lambda\le0.5$ to avoid the
aliasing (indeterminacy) the sinusoidal function, $A$ and $B$ are some constants.

Our objective is to perform a hypothesis test for detecting the periodicity,
$$\text{H}_{0}:\lambda=0\ \text{versus}\ \text{H}_{1}:\lambda>0.$$


### Fitting harmonic models with `ptest`

The function `fitHReg()` provides maximum likelihood estimations (MLE) for fitting the harmonic model. 
The default maximization of the likelihood is performed by the enumerative algorithm but an exact non-linear least square option is also provided.
Here is a brief example to do the estimations (See below). 

```{r Fit harmonic}
#simulate a harmonic series
set.seed(193)
#Non-Fourier frequency
z <- simHReg(n = 14, f=2/10, A = 2, B = 1, model="Gaussian",sig=1) 
fitHReg(z,algorithm="enumerative") 
fitHReg(z,algorithm = "exact") 
```

The recommended package boot may be used with fitHReg() to perform a bootstrap significance test for periodicity.

```{r boot test}
#Bootstrap the series under the null hypothesis: No periodicity 
set.seed(193)
library(boot)
LRStat <- function(d, i){    
  fitHReg(d[i],algorithm = "enumerative")$LRStat
}
bootResult <- boot(data = z, statistic = LRStat, R = 1000)
bootResult
#Compute the p-value
mean(bootResult$t>=fitHReg(z,algorithm="enumerative")$LRStat) #0.001

```

## Methods for detecting periodicity 

In order to detect whether the series is periodic or not, `ptest` provides two kinds of tests using the function `ptestReg()` and the function `ptestg()`. 
`ptestReg()` use likelihood ratio tests under the Gaussian or the Laplace assumptions, while `ptestg()` provides three methods based on the periodograms. 
Both methods are implemented with the RSR method implemented for efficiently obtaining accurate p-values.

### Likelihood ratio tests using `ptestReg()` 
`ptestReg()` performs two kinds of likelihood ratio tests based on two different assumptions (Gaussian or Laplace) and the p-value computation is efficiently improved by the RSR method.
Hence the p-value of the test can be instantly obtained without doing a Monte Carlo simulation and the precision is better than doing limited time of Monte Carlo simualtions in a personal computer. 
Here is a brief example.

```{r -2LLR} 
ptestReg(z,method = "LS") #Normal likelihood ratio test
ptestReg(z,method = "L1") #Laplace likelihood ratio test 
```

### Periodogram-based tests using `ptestg` 
The g statistic was initially proposed by [@fisher1929tests] and is used to test the periodicity in microarray data in [@Wichert01012004]. 
It is defined as a ratio of the maximum periodogram to the summation of all periodograms evaluated at the Fourier frequencies,
\begin{equation}
g=\frac{\max_{j}I(\lambda_{j})}{\sum_{j=1}^{m}I(\lambda_{j})},\label{eq:g-statistic}
\end{equation}
where $m$ is $(n-1)/2$ or $(n-2)/2$ according to $n$ is odd or
even and $I(\lambda_{j})$ is the periodogram evaluated at the Fourier frequencies $\lambda_{j}=j/n$, $j=1,...,[n/2]$
and $[x]$ denotes the integer part of x.

```{r Fisher}
ptestg(z,method="Fisher")
```

The other two periodogram-based tests are the modifications of the Fisher's g test. 
For the robust g test prposed by [@ahdesmaki2005robust], it replaces periodogram spectral estimator with an alternative rank-based robust estimator. 
The original robust g test staitstic is evaluated by enlarging twice the searching region of the fourier frequencies, but we also made an alternative option that the evaluated frequencies are $\cal{E}_{n}$.

```{r rank}
ptestg(z,method="robust")
#extend the frequency searching region to be En
ptestg(z,method="extendedRobust")
```

For the extended Fisher'g test statistic, it simply extends the Fisher's g test by enlarging the searching region of the frequency from the fourier frequencies to $\mathcal{E}_{n}$.

```{r extend}
ptestg(z,method="extended")
```


## Motivating Examples
Here we introduce some examples to stress the superiority of `ptest`.

### Efficiency with the response surface regression method
To see the efficiency of `ptest`, we compare the computation time of the robust g test in the `GeneCycle` package and the the robust g test in `ptest`. 
Here is the result on a single local machine with two cores Intel(R) Core(TM) i7-4500U CPU @ 1.80 GHz 2.40 GHz, Windows 10 64-bit system, and R 3.3.1.
Notice that here we only present the computation time of `GeneCycle` based on $10^5$ MC simulations, which is also the original set-up. 
But to achieve the same accuracy as the RSR method, it needs $200\times 10^5$ simulations, which is the simulation times used in the RSR method for the robust g in `ptest`.

```{r sim}
# Effiency Comparison for testing 10^5 series with length 20
set.seed(193)
X <- matrix(rnorm(20*10^5),nrow = 10)
```
```{r GeneCycle,eval=FALSE}
## MC method for the robust g test
library(GeneCycle)
# # The original robust g test performs 10^5 MC simualtions,
# # but we need 100*10^5 to achieve the same accuracy as 
# # the RSR method.
# # This process is time-consuming.
# t1 <- proc.time()
# RX <- robust.spectrum(X)
# pval1 <- robust.g.test(RX)
# t1 <- proc.time()-t1
# 
# t1 
##>    user  system elapsed 
##>  343.42    2.09  348.78
#unlink("g_pop_length_10.txt") # delete the external files 
```
hence for achieving the same accuracy as the below robust g test with the RSR method, it takes $348.78\times 200/3600 \ \text{hours} = 19.4\ \text{hours}$. 
Moreover, the memory needed is 99kb and hence the memory needed for the whole simulation will be almost 1GB. Fortunately, with the help of the RSR methods, those time-consuming simulations can be skipped. 

```{r RSR,eval=FALSE}
## RSR method for the robust g test
library(ptest)
t2 <- proc.time()
pval2 <- ptestg(X,method="robust",multiple = TRUE)
t2 <- proc.time()-t2

t2 
 #   user system  elapsed 
 # 208.97   0.17  218.42 
```

### Accuracy with the response surface regression method
To see the accuracy of computed p-values from the RSR method, the exact cumulative distribution function of 
the Fisher's g is compared to its cumulative distribution function derived from the RSR method.
For this experimental purpose, `ptestg()` provides an option for computing the p-value of 
the Fisher'g test with the RSR method.

```{r CDFFisherG, fig.show='hold', fig.width=7, fig.height=3.6}
#For a series of length 10, it is by the definition of the g statistic that 
#the range of g should be between 1/n and 1.
n <- 10
g <- seq(0.01,0.99,length.out = 200)

#Compute the CDF from the exact distribution and the RSR method
cdfExact <- pFisherg(g,n,method="exact")
cdfRSR <- pFisherg(g,n,method="RSR")


plot(g,cdfExact,col="red",type="l", cex=1.2,
     main = "Compare the exact CDF and the RSR CDF for Fisher's g (n=10)")
lines(g,cdfRSR,col="blue",type="b",pch=19,cex=0.3)
legend("topleft",legend = c("Exact","RSR"), 
       col = c("red","blue"), lty = c(1,NA),
       pch = c(NA, 19))

```

The absolute difference (in log 10 scale) between the exact distribution and the RSR/MC distribution is also shown. MC represents the emiprical distribution obtained from $10^5$ Monte Carlo simulations of the Fisher's g statistics. Usually this simulation step takes a long time (i.e., the robust g statistic), but since the g statistic is computed efficiently by the Fast Fourier transfomations, doing $10^5$ simulations is still acceptible. The figures shows that the RSR distribution is more accurate than the Monte Carlo method. 

```{r, fig.width=8.1, fig.height=4,fig.cap="Absolute difference between CDFs"}
set.seed(193)
zsim <- matrix(rnorm(10*10^5),nrow=10,ncol=10^5)
gSample <- ptestg(zsim,method="Fisher",multiple = TRUE)$obsStat
distMC <- ecdf(gSample)
cdfMC <- distMC(g)

#It is not of interest to compare thoes values which are less than 0.00001 
#or larger than 0.99999 as the 10^6 Monte Carlo simulations here does not 
#enable a reliable estimation for them.
pos <- which( (cdfExact>=0.00001)&(cdfExact<=0.99999) )
diff1 <- log(abs(cdfExact[pos]-cdfRSR[pos]),base = 10)
diff2 <- log(abs(cdfExact[pos]-cdfMC[pos]),base = 10)
comFig <- data.frame(cdfExact = cdfExact[pos],
                     diff=c(diff1,diff2),
                     type=rep(c("RSR v.s. Exact","MC v.s. Exact"),
                              each=length(diff1)))

library(lattice)
ans <- xyplot(diff ~ cdfExact | type, data=comFig,
       panel=function(x,y){
        panel.grid(h=4, v= 4,col=rgb(0.5,0.5,0.5,0.5))
        panel.xyplot(x, y, pch=16,cex=0.7)
       }, ylab="Absolute diference (log10)", xlab="g")
ans

```

```{r,include=FALSE}
# par(mfrow=c(1,2))
# plot(g[pos], diff1,
#      xlab = "g",
#      ylab = "absolute diference (log10)",
#      main = "The Exact v.s. RSR",
#      ylim = c(-7,-2))
# 
# plot(g[pos], diff2,
#      xlab = "g",
#      ylab = "absolute diference (log10)",
#      main = "The Exact v.s. MC",
#      ylim = c(-7,-2))
# par(mfrow=c(1,1))
```

### Might fail to detect non-fourier periodicities
Let's come back to the example of gene ORF06806. It is apparently periodic with period of about 8.4 hours, approximately with a non-Fourier frequency $1/8.4 = 0.1190$. But the Fisher's g failed to detect this periodicity.

```{r Motivation2, fig.cap="Gene ORF06806",fig.width=5,fig.height=3}
data(Cc)
x <- Cc[which(rownames(Cc)=="ORF06806"),]
plot(1:length(x),x,type="b",
     xlab="time",ylab="Gene expression")
```
```{r allTest}
ptestg(x,method="Fisher") #Fail to detect
ptestg(x,method="robust")
ptestg(x,method="extended")
ptestReg(x,method="LS")
ptestReg(x,method="L1")


X <- Cc[complete.cases(Cc),]
pvalue <- ptestg(t(X),method="Fisher",multiple = TRUE)

head(sort(pvalue$pvalue))

```

## Reference
